---
title: "NLP: Word embeddings"
output:
  html_notebook:
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

In this example, we are going to learn about an alternative method to encode 
text data known as ___word embeddings___. This is an incomplete tutorial on 
word embeddings but will at least give you the basic understanding on when, why,
and how we use them.

Learning objectives:

- What word embeddings are.
- The two main contexts that word embeddings are trained.
- When we should use word embeddings.
- How to train word embeddings for classification purposes.

# Requirements

```{r}
# Initialize package
library(keras)
library(fs)
library(tidyverse)
library(glue)
library(progress)

# helper functions we'll use to explore word embeddings
source("helper_functions.R")
```

# The "real" IMDB dataset

Keras provides a built-in IMBD dataset `dataset_imdb()`, which contains the
text of 25,000 movie reviews that have been classified as net positive or net
negative. However, we are going to use the original IMDB movie review files which
can be found at http://ai.stanford.edu/~amaas/data/sentiment/aclImdb_v1.tar.gz.
This tends to help students better understand the entire data prep required for
text.

You can find the download instructions [here](http://bit.ly/dl-rqmts). For those
in the workshop we have already downloaded this data for you.

```{r}
if (stringr::str_detect(here::here(), "conf-2020-user")) {
  imdb_dir <- "/home/conf-2020-user/data/imdb"
} else {
  imdb_dir <- here::here("materials", "data", "imdb")
}
fs::dir_tree(imdb_dir, type = "directory")
```

You can see the data have already been separated into test vs training sets and 
positive vs negative sets. The actual reviews are contained in individual .txt
files. We can use this structure to our advantage - the below iterates over each
review and 

1. creates the path to each individual review file,
2. creats a label based on the "neg" or "pos" folder the review is in,
3. and saves the output as a data frame with each review on an individual row.

```{r}
training_files <- file.path(imdb_dir, "train") %>%
  dir_ls(type = "directory") %>%
  map(dir_ls) %>%
  set_names(basename) %>%
  plyr::ldply(data_frame) %>%
  set_names(c("label", "path"))

training_files
```

We can see our response observations are balanced:

```{r}
count(training_files, label)
```

We can now iterate over each row and

1. save the label in a `labels` vector,
2. import the movie review and
3. save in a `texts` vector.

```{r}
obs <- nrow(training_files)
labels <- vector(mode = "integer", length = obs)
texts <- vector(mode = "character", length = obs)

# this just allows us to track progress of our loop
pb <- progress_bar$new(total = obs, width = 60)

for (file in seq_len(obs)) {
  pb$tick()
  
  label <- training_files[[file, "label"]]
  path <- training_files[[file, "path"]]
  
  labels[file] <- ifelse(label == "neg", 0, 1)
  texts[file] <- readChar(path, nchars = file.size(path)) 
  
}
```

We now have two vectors, one consisting of the labels...

```{r}
table(labels)
```

and the other holding each review.

```{r}
texts[1]
```

# Exploratory text analysis

A little exploratory analysis will show us the total number of unique words
across our corpus and the average length of each review. Its good to know the
word count distribution of your text as later on we'll make a decision of how
many words to keep.

```{r, fig.height=3.5}
text_df <- texts %>%
  tibble(.name_repair = ~ "text") %>%
  mutate(text_length = str_count(text, "\\w+"))

unique_words <- text_df %>%
  tidytext::unnest_tokens(word, text) %>%
  pull(word) %>%
  n_distinct()

avg_review_length <- median(text_df$text_length, na.rm = TRUE)
  
ggplot(text_df, aes(text_length)) +
  geom_histogram(bins = 100, fill = "grey70", color = "grey40") +
  geom_vline(xintercept = avg_review_length, color = "red", lty = "dashed") +
  scale_x_log10("# words") +
  ggtitle(glue("Median review length is {avg_review_length} words"),
          subtitle = glue("Total number of unique words is {unique_words}"))
```


# Word embeddings for language modeling

Word embeddings are designed to encode general semantic relationships which can
serve two principle purposes. The first is for ___language modeling___ which 
aims to encode words for the purpose of predicting synonyms, sentence completion, 
and word relationships. [ℹ️](http://bit.ly/dl-06)

Although we are not focusing on word embeddings for this purpose, I have written
a couple helper functions to train word embeddings for this purpose. See the
code behind these helper functions [here](https://bit.ly/32HCP1G).

```{r}
# clean up text and compute word embeddings
clean_text <- tolower(texts) %>%
  str_replace_all(pattern = "[[:punct:] ]+", replacement = " ") %>%
  str_trim()

word_embeddings <- get_embeddings(clean_text)
```

Explore your own words!

```{r}
# find words with similar embeddings
get_similar_words("horrible", word_embeddings)
```


# Word embeddings for classification

The other principle purpose for word embeddings is to encode text for 
classification reasons. In this case, we train the word embeddings to take on 
weights that optimize the classification loss function. [ℹ️](http://bit.ly/dl-06#13)

## Prepare data

Our response variable `labels` is already a tensor; however, we still need to
preprocess our text features. To do so we:

1. Specify how many words we want to include. This example uses the 10,000 words
   with the highest usage (frequency).
2. Create a `text_tokenizer` object which defines how we want to preprocess the
   text (i.e. convert to lowercase, remove punctuation, token splitting 
   characters). For the most part, the defaults are sufficient.
3. Apply the tokenizer to our text with `fit_text_tokenizer`. This results in an
   object with many details of our corpus (i.e. word counts, word index).

```{r}
top_n_words <- 10000

tokenizer <- text_tokenizer(num_words = top_n_words) %>% 
  fit_text_tokenizer(texts)

names(tokenizer)
```

We have now tokenized our reviews. We are considering 10,000 of 88,582 total
unique words. The most common words include:
     
```{r}
head(tokenizer$word_index)
```

Next, we extract our vectorized review data as a list. Each review is encoded as
a sequence of word indexes (integers).

```{r}
sequences <- texts_to_sequences(tokenizer, texts)

# The vectorized first instance:
sequences[[1]]
```

We can map the integer values back to the word index. The integer number
corresponds to the position in the word count list and the name of the vector is
the actual word.

```{r}
paste(unlist(tokenizer$index_word)[sequences[[1]]] , collapse = " ")
```

We can see how our tokenizer converted our original text to a cleaned up 
version:

```{r} 
cat(crayon::blue("Original text:\n"))
texts[[1]]

cat(crayon::blue("\nRevised text:\n"))
paste(unlist(tokenizer$index_word)[sequences[[1]]] , collapse = " ")
```

Next, since each review is a different length, we need to limit ourselves to a
certain number of words so that all our features (reviews) are the same length.
This should be viewed as a tuning parameter. 

__Tip__: I typically start with values around the 50% (median) but then explore
values that represent 25% & 75% percentile of the word distribution when tuning.

Note (`?pad_sequences`):

* Any reviews that are shorter than this length will be padded.
* Any reviews that are longer than this length will be truncated.

```{r}
max_len <- 150
features <- pad_sequences(sequences, maxlen = max_len)
```

Since this review includes less than 150 words from our word index of 10K most
frequent words, it pads the front-end with zeros (see `?pad_sequences()` for
alternative padding options).

```{r}
features[1,]
```

So, in essence, we have created an input that is a numeric representation of
this:

```{r}
features[1,] %>% 
  map_chr(~ ifelse(.x == 0, "<pad>", unlist(tokenizer$index_word[.x]))) %>%
  cat()
```

### Your Turn!

Check out different reviews and see how we have transformed the data. Remove 
`eval=FALSE` to run.

```{r, eval=FALSE}
# use review number (i.e. 2, 10, 150)
which_review <- ____
  
cat(crayon::blue("Original text:\n"))
texts[[which_review ]]

cat(crayon::blue("\nRevised text:\n"))
paste(unlist(tokenizer$index_word)[features[which_review ,]] , collapse = " ")

cat(crayon::blue("\nEncoded text:\n"))
features[which_review,] %>% 
  map_chr(~ ifelse(.x == 0, "<pad>", unlist(tokenizer$index_word[.x]))) %>%
  cat()
```


Our data is now preprocessed! We have `r nrow(features)` observations and 
`r ncol(features)` features. Our `features` data is a matrix where each row is
a single observation and each column represents the words in the review in the
order that they appear.

```{r}
dim(features)
length(labels)
```


## Model training

To train our model we will use the `validation_split` procedure within `fit`. 
Remember, this takes the last XX% of our data to be used as our validation set. 
But if you recall, our data was organized in neg and pos folders so we should 
randomize our data to make sure our validation set doesn't end up being all 
positive or negative reviews!

```{r}
set.seed(123)
index <- sample(1:nrow(features))

x_train <- features[index, ]
y_train <- labels[index]
```

To create our network architecture that includes word embeddings, we need to 
include two things:

1. `layer_embedding` layer that creates the embeddings,
2. `layer_flatten` to flatten our embeddings to a 2D tensor for our densely 
    connected portion of our model
    
__Note__:

* `input_dim` & `input_length` are considered pre-processing hyperparameters.
* `output_dim` is our word embeddings hyperparameter.
* They all have interaction effects

```{r}
model <- keras_model_sequential() %>%
  layer_embedding(
    input_dim = top_n_words,   # number of words we are considering
    input_length = max_len,    # length that we have set each review to
    output_dim = 32            # length of our word embeddings
    ) %>%  
  layer_flatten() %>% 
  layer_dense(units = 1, activation = "sigmoid")

summary(model)
```

The rest of our modeling procedure follows the same protocols that you've seen 
in the other modules.

```{r}
model %>% compile(
  optimizer = "rmsprop",
  loss = "binary_crossentropy",
  metrics = "accuracy"
)

history <- model %>% fit(
  x_train, y_train,
  epochs = 10,
  batch_size = 32,
  validation_split = 0.2
)
```

```{r}
best_epoch <- which.min(history$metrics$val_loss)
best_loss <- history$metrics$val_loss[best_epoch] %>% round(3)
best_acc <- history$metrics$val_accuracy[best_epoch] %>% round(3)

glue("Our optimal loss is {best_loss} with an accuracy of {best_acc}")
```


```{r, message=FALSE}
plot(history)
```

## YOUR TURN! (5min)

Spend a few minutes adjusting this model and see how it impacts performance. You
may want to test:

- Does increasing and decreasing the word embedding dimension (`output_dim`)
  impacts performance?
- How does the learning rate impact performance?
- You may have noticed that we didn't add any additional hidden layers to the
  densely connected portion of our model.  Does adding 1 or 2 more hidden layers
  improve performance?

```{r, eval=FALSE}
yourturn_model <- keras_model_sequential() %>%
  layer_embedding(
    input_dim = _____,  
    input_length = _____,   
    output_dim = _____           
    ) %>%  
  layer_flatten() %>% 
  layer_dense(units = ____, activation = ____) %>%
  layer_dense(units = 1, activation = "sigmoid")

yourturn_model %>% compile(
  optimizer = _____,
  loss = "binary_crossentropy",
  metrics = "accuracy"
)

yourturn_results <- yourturn_model %>% fit(
  x_train, y_train,
  epochs = 10,
  batch_size = 32,
  validation_split = 0.2
)
```

# Comparing embeddings

Recall that the word embeddings we found for natural language modeling created 
results like:

```{r}
# natural language modeling embeddings
get_similar_words("horrible", word_embeddings)
```

However, embeddings we find for classification tasks are not always so clean and 
intuitive. We can get the word embeddings from our classification model with:

```{r}
wts <- get_weights(model)  # returns a list with each layers weights
embedding_wts <- wts[[1]]  # get the first layer's weights
```

The following just does some bookkeeping to extract the applicable words and 
assign them as row names to the embedding matrix.

```{r}
words <- tokenizer$word_index %>% 
  as_tibble() %>% 
  pivot_longer(everything(), names_to = "word", values_to = "id") %>%
  filter(id <= tokenizer$num_words) %>%
  arrange(id)

row.names(embedding_wts) <- words$word
```

The following is one of the custom functions you imported from the 
[helper_functions.R](https://bit.ly/32HCP1G) file. You can see the word 
embeddings that most closely align to a given word are not as intuitive as those
produced from the natural language model. However, these are the embeddings that
optimized for the classification procedure at hand.

```{r}
similar_classification_words("horrible", embedding_wts)
```
 
Here's a handy sequence of code that uses the [t-SNE](https://bit.ly/2rDk6rs) 
methodology to visualize nearest neighbor word embeddings.

```{r, fig.width=10, fig.height=6}
# plotting too many words makes the output hard to read
n_words_to_plot <- 1000

tsne <- Rtsne::Rtsne(
  X = embedding_wts[1:n_words_to_plot,], 
  perplexity = 100, 
  pca = FALSE
  )

p <- tsne$Y %>%
  as.data.frame() %>%
  mutate(word = row.names(embedding_wts)[1:n_words_to_plot]) %>%
  ggplot(aes(x = V1, y = V2, label = word)) + 
  geom_text(size = 3)

plotly::ggplotly(p)
```
 
# Key takeaways
 
* Word embeddings
   - Commonly used for language modeling and prediction tasks.
   - Provide a dense, relationship rich vector representation of words.
   - We can measure similarity with cosine similarity (other distance functions
      are also used).
   - See http://bit.ly/dl-06#12 for list of resources to go deeping into word
     embeddings.

* Data prep
   - Use `text_tokenizer()` to define the text preprocessing we desire (defaults
     are good).
   - Use `fit_text_tokenizer()` to preprocess text (i.e. remove punctuation,
     standardize to lowercase).
   - Use `texts_to_sequences()` to convert standardized text to numeric
     representation of word index.
   - Use `pad_sequences()` to make all text sequences the same length. This
     length can be adjusted to improve performance.

* Model training
   - Fit word embeddings with `layer_embedding()`.
   - Tune `output_dim`ension of the embeddings.
   - Apply `layer_flatten()` to convert embeddings to a 2D tensor and use
     `layer_dense()` for classification (there are alternatives which we will
     cover later).

[🏠](https://github.com/rstudio-conf-2020/dl-keras-tf)