---
title: "NLP: Transfer learning with GloVe word embeddings"
output:
  html_notebook:
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

In this example, we are going to learn how to apply pre-trained word embeddings.
This can be useful when you have a very small dataset; too small to actually
learn the embeddings from the data itself. However, pre-trained word embeddings
for regression and classification predictive purposes rarely perform as well as
learning the word embeddings from the data itself. This will become obvious in
this example.

Learning objectives:

- How to prepare pre-trained word embeddings
- How to apply pre-trained word embeddings

# Requirements

```{r}
library(keras)     # deep learning modeling
library(tidyverse) # various data wrangling & visualization tasks
library(progress)  # provides progress bar for status updates during long loops
library(glue)      # easy print statements
```

# Preprocess IMDB data

To minimize notebook clutter, I have extracted the code to preprocess the
training data into a separate script. This script uses the same downloaded IMDB
movie reviews that the previous module used and preprocesses it in the same
way.

```{r, warning=FALSE}
source("prepare_imdb.R")
```

As a result, we have a few key pieces of information that we will use in
downstream modeling (i.e. word index, features, max sequence length).

```{r}
ls()
```

# Prepare GloVe pre-trained word embeddings

We are going to use the pre-trained GloVe word embeddings which can be downloaded
[here](https://nlp.stanford.edu/projects/glove/). For this example, we downloaded
the glove.6B.zip file that contains 400K words and their associated word
embeddings. Here, we'll use the 100 dimension word embeddings which has already
been saved for you in the data directory. 

___See the [requirements set-up notebook](http://bit.ly/dl-rqmts) for download instructions.___

```{r}
# get path to downloaded data
if (stringr::str_detect(here::here(), "conf-2020-user")) {
  path <- "/home/conf-2020-user/data/glove/glove.6B.100d.txt"
} else {
  path <- here::here("materials", "data", "glove", "glove.6B.100d.txt")
}

glove_wts <- data.table::fread(path, quote = "", data.table = FALSE) %>% 
  as_tibble()

dim(glove_wts)
```

Our imported data frame has the associated word (or grammatical symbol), and the
100 weights associated to its representative vector.

```{r}
head(glove_wts)
```

However, pre-trained models are typically trained on entirely different data (or
vocabulary). Consequently, they do not always capture all words present in our
dataset. The following illustrates that...

```{r}
applicable_index <- total_word_index[total_word_index <= top_n_words]
applicable_words <- names(applicable_index)

available_wts <- glove_wts %>%
  filter(V1 %in% applicable_words) %>% 
  pull(V1)

diff <- length(applicable_words) - length(available_wts)

glue("There are {diff} words in our IMDB data that are not represented in GloVe")
```

We need to create our own embeddings matrix with all applicable words
represented. When doing so, we want to create the matrix in order of our word
index so the embeddings are properly aligned. To do so, we will create an empty
matrix to fill.

```{r}
# required dimensions of our embedding matrix
num_words_used <- length(applicable_words)
embedding_dim <- ncol(glove_wts) - 1

# create empty matrix
embedding_matrix <- matrix(0, nrow = num_words_used, ncol = embedding_dim)
row.names(embedding_matrix) <- applicable_words

# First 10 rows & columns of our empty matrix
embedding_matrix[1:10, 1:10]
```

To fill our embedding matrix, we loop through the GloVe weights, get the
available embeddings, and add to our empty embedding matrix so that they align 
with the word index order.  If the word does not exist in the pretrained word
embeddings then we make the embedding values 0.

**Note: this takes a little less than 2 minutes to process.**

```{r}
# this just allows us to track progress of our loop
pb <- progress_bar$new(total = num_words_used)

for (word in applicable_words) {
  # track progress
  pb$tick()
  
  # get embeddings for a given word
  embeddings <- glove_wts %>%
    filter(V1 == word) %>%
    select(-V1) %>% 
    as.numeric()
  
  # if embeddings don't exist create a vector of all zeros
  if (all(is.na(embeddings))) {
    embeddings <- vector("numeric", embedding_dim)
  }
  
  # add embeddings to appropriate location in matrix
  embedding_matrix[word, ] <- embeddings
}
```

```{r}
embedding_matrix[1:10, 1:8]
```

Now we're ready to train our model.

# Create and train our model

## Define a model

We will be using the same model architecture as we did in the IMDB word 
embeddings [notebook](http://bit.ly/dl-imdb-embeddings). The `top_n_words` and
`max_len` objects come from the `IMDB `source("prepare_imdb.R")` code chunk.

```{r}
model <- keras_model_sequential() %>% 
  layer_embedding(input_dim = top_n_words, 
                  output_dim = embedding_dim, 
                  input_length = max_len) %>% 
  layer_flatten() %>% 
  layer_dense(units = 1, activation = "sigmoid")

summary(model)
```

## Load the GloVe embeddings in the model

To set the weights of our embedding layer to our pretrained embedding matrix, 
we:

1. access our first layer,
2. set the weights by supplying our embedding matrix,
3. freeze the weights so they are not adjusted when we train our model.

```{r}
get_layer(model, index = 1) %>% 
  set_weights(list(embedding_matrix)) %>% 
  freeze_weights()
```

## Train and evaluate

Let's compile our model and train it.

```{r, echo=TRUE, results='hide'}
model %>% compile(
  optimizer = optimizer_rmsprop(lr = 0.0001),
  loss = "binary_crossentropy",
  metrics = c("acc")
)

history <- model %>% fit(
  features, labels,
  epochs = 20,
  batch_size = 32,
  validation_split = 0.2,
  callbacks = list(
    callback_early_stopping(patience = 5),
    callback_reduce_lr_on_plateau(patience = 2)
    )
)
```

Our best performance is less than stellar!

```{r}
best_epoch <- which(history$metrics$val_loss == min(history$metrics$val_loss))
loss <- history$metrics$val_loss[best_epoch] %>% round(3)
acc <- history$metrics$val_acc[best_epoch] %>% round(3)

glue("The best epoch had a loss of {loss} and accuracy of {acc}")
```

# When to use pre-trained models?

Recall that word embeddings are often trained for language models
[ℹ️](http://bit.ly/dl-06#2). This means the embeddings are designed to predict
the optimal word within a phrase. This rarely generalizes well to using word
embeddings for normal regression and classification modeling. 

Things to consider:

- When you have very little text data, you often cannot produce reliable
  embeddings. Pre-trained embeddings can sometimes provide performance
  improvements in these situations.

- Even better, when you have very little data, find alternative data sources
  that aligns with your (i.e. use Amazon product reviews as an alternative
  data source to your company's product reviews). You can then train word
  embeddings on this alternative data source and use those embeddings for your
  own problem.

- Some language models (i.e. [ULMFiT](http://nlp.fast.ai/category/classification.html))
  can be transferred for classification purposes much like VGG16, Resnet50, etc.
  within the CNNs context. Consequently, you can use these models, freeze and
  unfreeze layer weights, and re-train only part of the model to be specific to
  your task.

[🏠](https://github.com/rstudio-conf-2020/dl-keras-tf)