---
title: "Case Study 1: Single-label, Multi-class Classification with MNIST"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

# Initialize package
# install.packages(keras)
library(keras)
library(tidyverse)
```

# Install tensorflow 

It's only necessary to run this once. 

```{r install, eval = F}
# for GPU
# install_keras(tensorflow = "gpu")

# or CPU:
# install_keras() # for cpu
```

## Part 1: Data Preparation

## Obtain data

```{r data}
mnist <- dataset_mnist()
c(c(train_images, train_labels), c(test_images, test_labels)) %<-% mnist

load("MNISTCollection.RData")

```

## View images:

```{r plotdigit1, fig.show = 'hold', out.width = '25%'}
  image <- 1
  plot(as.raster(train_images_original[image,,], max = 255), main = "original")
  plot(as.raster(train_images_rearrange[image,,], max = 255), main = "rearrange")
  plot(as.raster(train_images_scramble[image,,], max = 255), main = "scramble")
  plot(as.raster(train_images_shift[image,,], max = 255), main = "shift")
```

## Visualizing the input

Let's look at for example what all the 1's would look like:

```{r plot numbers, cache = T}

train_images_original_reshaped <- array_reshape(train_images_original, c(60000, 28 * 28))/255
train_images_original_reshaped <- as_tibble(train_images_original_reshaped)
train_images_original_reshaped$type <- 1


train_images_rearrange_reshaped <- array_reshape(train_images_rearrange, c(60000, 28 * 28))/255
train_images_rearrange_reshaped <- as_tibble(train_images_rearrange_reshaped)
train_images_rearrange_reshaped$type <- 2

train_images_scramble_reshaped <- array_reshape(train_images_scramble, c(60000, 28 * 28))/255
train_images_scramble_reshaped <- as_tibble(train_images_scramble_reshaped)
train_images_scramble_reshaped$type <- 3

train_images_shift_reshaped <- array_reshape(train_images_shift, c(60000, 28 * 28))/255
train_images_shift_reshaped <- as_tibble(train_images_shift_reshaped)
train_images_shift_reshaped$type <- 4


rbind(train_images_original_reshaped,
      train_images_rearrange_reshaped,
      train_images_scramble_reshaped,
      train_images_shift_reshaped) %>%
  as_tibble() %>% 
  mutate(id = rep(as.factor(train_labels), 4)) %>%
  group_by(type, id) %>% 
  summarise_if(is.numeric, sum) %>% 
  ungroup() %>% 
  gather(key, value, -c(type, id)) %>% 
  arrange(type, id) %>% 
  mutate(x = rep(rep(rep(1:28, 28), 10),4),
         y = rep(rep(rep(28:1, each = 28),10),4)) -> NEWDF

library(RColorBrewer)
ggplot(NEWDF, aes(x, y, fill = value)) +
  geom_raster() +
  coord_fixed() +
  scale_fill_gradientn(colours = c("black", "white")) +
  # scale_fill_gradientn(colours = (brewer.pal(9, "Reds"))) +
  facet_grid(type ~ id) +
  theme_void() +
  theme(legend.position = "none",
        strip.text = element_blank())

# ggsave("compiledNumbers_bw.png", height = 6, width = 18, units = "in")

```


## Normalize data:

### Training set

```{r normImages, echo = FALSE}

# # Before making a list:
# train_images_original[1,,]
# train_images_rearrange[1,,]
# train_images_scramble[1,,]
# train_images_shift[1,,]

# Make a list:
train_images <- list(train_images_original,
                     train_images_rearrange,
                     train_images_scramble,
                     train_images_shift)

# # After making a list:
# train_images[[1]][1,,]
# train_images[[2]][1,,]
# train_images[[3]][1,,]
# train_images[[4]][1,,]

# Normalize
train_images[[1]] <- array_reshape(train_images[[1]], c(60000, 28 * 28))/255
train_images[[2]] <- array_reshape(train_images[[2]], c(60000, 28 * 28))/255
train_images[[3]] <- array_reshape(train_images[[3]], c(60000, 28 * 28))/255
train_images[[4]] <- array_reshape(train_images[[4]], c(60000, 28 * 28))/255

# # After Normalizing:
# train_images[[1]][1,]
# train_images[[2]][1,]
# train_images[[3]][1,]
# train_images[[4]][1,]

```

### Test set

```{r normTest, echo = FALSE}


# # Before making a list:
# test_images_original[1,,]
# test_images_rearrange[1,,]
# test_images_scramble[1,,]
# test_images_shift[1,,]

# Make a list:
test_images <- list(test_images_original,
                    test_images_rearrange,
                    test_images_scramble,
                    test_images_shift)

# # After making a list:
# test_images[[1]][1,,]
# test_images[[2]][1,,]
# test_images[[3]][1,,]
# test_images[[4]][1,,]

# Normalize
test_images[[1]] <- array_reshape(test_images[[1]], c(10000, 28 * 28))/255
test_images[[2]] <- array_reshape(test_images[[2]], c(10000, 28 * 28))/255
test_images[[3]] <- array_reshape(test_images[[3]], c(10000, 28 * 28))/255
test_images[[4]] <- array_reshape(test_images[[4]], c(10000, 28 * 28))/255

# # After Normalizing:
# test_images[[1]][1,]
# test_images[[2]][1,]
# test_images[[3]][1,]
# test_images[[4]][1,]


```


## Prepare labels:

```{r prepLabels}
train_labels <- to_categorical(train_labels)
test_labels <- to_categorical(test_labels)
```

# Part 2: Define Network

## Define & Compile the network

```{r architecture}
makeNetwork <- function() {
  keras_model_sequential() %>%
    layer_dense(units = 512, activation = "relu", input_shape = c(28 * 28)) %>%
    layer_dense(units = 10, activation = "softmax") %>%
    compile(
      optimizer = "rmsprop",
      loss = "categorical_crossentropy",
      metrics = c("accuracy")
      )
}
```

## Train the model

### Model 1

```{r train1}
network1 <- makeNetwork()
# summary(network1)

network1 %>%
  fit(train_images[[1]], train_labels, epochs = 5, batch_size = 128) -> history1
```

### Model 2

```{r train2}
network2 <- makeNetwork()
# summary(network2)

network2 %>%
  fit(train_images[[2]], train_labels, epochs = 5, batch_size = 128) -> history2
```

### Model 3

```{r train3}
network3 <- makeNetwork()
# summary(network3)

network3 %>%
  fit(train_images[[3]], train_labels, epochs = 5, batch_size = 128) -> history3
```

### Model 4

```{r train4}
network4 <- makeNetwork()
# summary(network4)

network4 %>%
  fit(train_images[[4]], train_labels, epochs = 5, batch_size = 128) -> history4
```


# Part 3: Check output
## Plot all histories

```{r plotTraining1}
plot(history1)

AllHistory <- rbind(
  data.frame(history1,
             id = 1),
  data.frame(history2,
             id = 2),
  data.frame(history3,
             id = 3),
  data.frame(history4,
             id = 4)
  )

levels(AllHistory$metric) <- c("Accuracy", "Loss")

library(ggplot2)
ggplot(AllHistory, aes(epoch, value, col = factor(id))) +
  geom_line(size = 1, alpha = 0.65, position = position_dodge(0.2)) +
  facet_wrap(~metric, scales = "free", nrow = 2) +
  scale_color_brewer("Series", 
                     palette = "Dark2",
                     labels = c("Original",
                                "Rearranged\nquadrants",
                                "Rearranged\nrows and columns",
                                "Random shift")) +
  labs(x = "Epochs", y = "Value") +
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text = element_text(color = "black"))
  
# ggsave("training_results.pdf", height = 6, width = 10, units = "in")

  
```

```{r}
AllHistory %>% 
  spread(metric, value) %>% 
  filter(epoch == 5) %>% 
  knitr::kable() 
```


## Metrics

```{r metrics1}

library(tidyr)
library(dplyr)
data.frame(X1 = unlist(network1 %>% evaluate(test_images[[1]], test_labels)),
           X2 = unlist(network2 %>% evaluate(test_images[[2]], test_labels)),
           X3 = unlist(network3 %>% evaluate(test_images[[3]], test_labels)),
           X4 = unlist(network4 %>% evaluate(test_images[[4]], test_labels))) %>% 
  mutate(metric = row.names(.)) %>% 
  gather(key, value, -metric) -> results_test 

results_test$key <- factor(results_test$key, labels = c("Original",
                                                        "Rearranged\nquadrants",
                                                        "Rearranged\nrows and columns",
                                                        "Random shift"))
results_test$metric <- factor(results_test$metric, labels = c("Accuracy", "Loss"))

results_test %>% 
  ggplot(aes(col = key, y = value, x = metric)) +
  geom_point(alpha = 0.65, position = position_dodge(0.2)) +
  scale_color_brewer("Series", 
                     palette = "Dark2") +
  labs(x = "Metric", y = "Value") +
  theme_classic() +
  theme(axis.line.x = element_blank(),
        axis.text = element_text(color = "black"))

ggsave("test_results.pdf", height = 10, width = 6, units = "in")


```
