---
title: "Computer Vision: Visualizing Cats and Dogs"
author: "Rick Scavetta"
output: html_document
---

# Data sources

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

# Initialize package
library(keras)

# define the directories:
source("def_dir.R")
```

# Obtain a pre-trained networks

```{r getModel}

model <- load_model_hdf5("cats_and_dogs_small_2.h5")
model

```

## view

```{r getIMG}

img_path <- "~/Downloads/cats_and_dogs_small/test/cats/cat.1700.jpg"
img <- image_load(img_path, target_size = c(150, 150))
img_tensor <- image_to_array(img)
img_tensor <- array_reshape(img_tensor, c(1, 150, 150, 3))
img_tensor <- img_tensor / 255

# Preprocesses the image into a 4D tensor
   dim(img_tensor)
   
```

# plot image:

```{r vizIMG}

plot(as.raster(img_tensor[1,,,]))

```

# Instantiating a model from an input tensor and a list of output tensors

```{r act0}

layer_outputs <- lapply(model$layers[1:8], function(layer) layer$output)
activation_model <- keras_model(inputs = model$input, outputs = layer_outputs)

```

# Running the model in predict mode

```{r act1}

activations <- activation_model %>% predict(img_tensor)

```

this is the activation of the first convolution layer for the cat image input:

```{r act2}

first_layer_activation <- activations[[1]]
dim(first_layer_activation)

```

# Function to plot a channel

```{r vizFUN}

plot_channel <- function(channel) {
  rotate <- function(x) t(apply(x, 2, rev))
  image(rotate(channel), axes = FALSE, asp = 1,
        col = terrain.colors(12))
}

```

# Plotting the second channel

```{r viz2}

plot_channel(first_layer_activation[1,,,2])

```

# Visualizing the seventh channel

```{r viz7}

plot_channel(first_layer_activation[1,,,7])

```

# Visualizing every channel in every intermediate activation

```{r vizAll}

image_size <- 58
images_per_row <- 16

for (i in 1:8) {
  layer_activation <- activations[[i]]
  layer_name <- model$layers[[i]]$name
  n_features <- dim(layer_activation)[[4]]
  n_cols <- n_features %/% images_per_row
  png(paste0("cat_activations_", i, "_", layer_name, ".png"),
      width = image_size * images_per_row,
      height = image_size * n_cols)
  op <- par(mfrow = c(n_cols, images_per_row), mai = rep_len(0.02, 4))
  for (col in 0:(n_cols-1)) {
    for (row in 0:(images_per_row-1)) {
      channel_image <- layer_activation[1,,,(col*images_per_row) + row + 1]
      plot_channel(channel_image)
    }
  }
  par(op)
  dev.off() }

```

