---
title: "Computer Vision: Optimization"
author: "Rick Scavetta"
output: html_document
---

# Data sources

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

# Initialize package
library(keras)

# define the directories:
source("def_dir.R")
```

# Obtain a pre-trained convnet base

```{r getConv}
conv_base <- application_vgg16(
  weights = "imagenet",
  include_top = FALSE,
  input_shape = c(150, 150, 3)
)

```

```{r summaryConv}
conv_base
```

# Extracting features using the pretrained convolutional base

```{r dataGen0}

datagen <- image_data_generator(rescale = 1/255)
batch_size <- 20

extract_features <- function(directory, sample_count) {
  features <- array(0, dim = c(sample_count, 4, 4, 512))
  labels <- array(0, dim = c(sample_count))
  generator <- flow_images_from_directory(
    directory = directory,
    generator = datagen,
    target_size = c(150, 150),
    batch_size = batch_size,
    class_mode = "binary"
  )
  i <- 0
  while(TRUE) {
    batch <- generator_next(generator)
    inputs_batch <- batch[[1]]
    labels_batch <- batch[[2]]
    features_batch <- conv_base %>% predict(inputs_batch)
    index_range <- ((i * batch_size)+1):((i + 1) * batch_size)
    features[index_range,,,] <- features_batch
    labels[index_range] <- labels_batch
    i <- i + 1
    if (i * batch_size >= sample_count)
      break }
  list(
    features = features,
    labels = labels
  ) }


train <- extract_features(train_dir, 2000)
validation <- extract_features(validation_dir, 1000)
test <- extract_features(test_dir, 1000)

# train <- extract_features(paste0("/home/rstudio-user",sub("~","",train_dir)), 2000)
# validation <- extract_features(paste0("/home/rstudio-user",sub("~","",validation_dir)), 1000)
# test <- extract_features(paste0("/home/rstudio-user",sub("~","",test_dir)), 1000)

```

# Reshape features

```{r reshape}
reshape_features <- function(features) {
  array_reshape(features, dim = c(nrow(features), 4 * 4 * 512))
}
train$features <- reshape_features(train$features)
validation$features <- reshape_features(validation$features)
test$features <- reshape_features(test$features)
```

# Define model

Here we only need the densely connected classifier

```{r denseModel}

model <- keras_model_sequential() %>%
  layer_dense(units = 256, activation = "relu", input_shape = 4 * 4 * 512) %>%
  layer_dropout(rate = 0.5) %>%
  layer_dense(units = 1, activation = "sigmoid")

model

```

Compile and train:

```{r modelTrain}

model %>% compile(
  optimizer = optimizer_rmsprop(lr = 2e-5),
  loss = "binary_crossentropy",
  metrics = c("accuracy")
)

history1 <- model %>% fit(
  train$features, train$labels,
  epochs = 30,
  batch_size = 20,
  validation_data = list(validation$features, validation$labels)
)

```

# Plot results:

```{r hist1}
plot(history1)
```

# Combining a densely-connected neural network with the convolutional base

```{r convModel}

model <- keras_model_sequential() %>%
  conv_base %>%
  layer_flatten() %>%
  layer_dense(units = 256, activation = "relu") %>%
  layer_dense(units = 1, activation = "sigmoid")

model

```

## Freezing

```{r freezeParam}
cat(length(model$trainable_weights), " trainable weights before freezing.\n")

freeze_weights(conv_base)

cat(length(model$trainable_weights), " trainable weights before freezing.\n")

```

# Training the model end-to-end with a frozen convolutional base

```{r datagen1}

train_datagen = image_data_generator(
  rescale = 1/255,
  rotation_range = 40,
  width_shift_range = 0.2,
  height_shift_range = 0.2,
  shear_range = 0.2,
  zoom_range = 0.2,
  horizontal_flip = TRUE,
  fill_mode = "nearest"
)

test_datagen <-
  image_data_generator(rescale = 1/255)

train_generator <- flow_images_from_directory(
  train_dir,
  train_datagen,
  target_size = c(150, 150),
  batch_size = 20,
  class_mode = "binary"
)

validation_generator <- flow_images_from_directory(
  validation_dir,
  test_datagen,
  target_size = c(150, 150),
  batch_size = 20,
  class_mode = "binary"
)

model %>% compile(
  loss = "binary_crossentropy",
  optimizer = optimizer_rmsprop(lr = 2e-5),
  metrics = c("accuracy")
)

history2 <- model %>% fit_generator(
  train_generator,
  steps_per_epoch = 100,
  epochs = 30,
  validation_data = validation_generator,
  validation_steps = 50
)

```

# View history

```{r hist2s}
plot(history2)
```
