---
title: "Installing Requirements"
output: github_document
---

This notebook will help you install all the required packages and data sets used
throughout this workshop.

# Packages

Although this training focuses on the __keras__ package, we will use several
additional packages throughout our modules. At the beginning of each module I
state the primary purpose of the packages used so if some of these packages are
not familiar, that is okay.

Run the following code chunk to install packages not already installed. Within
the RStudio workshop, this has already been taken care of for you.

```{r}
pkgs <- c(
  "AmesHousing",
  "caret",
  "crayon",
  "data.table",
  "fs",
  "glue",
  "here",
  "jpeg",
  "keras",
  "plotly",
  "recipes",
  "ROCR",
  "Rtsne",
  "rsample",
  "testthat",
  "text2vec",
  "tidyverse"
)


missing_pkgs <- pkgs[!(pkgs %in% installed.packages()[, "Package"])]

if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs)
}
```

Depending on how often you update your packages you may also want to see if any
of these packages that you had installed previously need to be updated to more
recent versions. You can run the following to make necessary updates.

```{r}
update.packages(oldPkgs = pkgs, ask = FALSE, repos = "https://cran.rstudio.com/")
```

# Datasets

We will use a variety of datasets throughout this workshop; many of which are
too large to hold within the github repository. The following will help you 
download all necessary datasets and store them in the `\data` folder. For those
attending the RStudio workshop, this has already been taken care of for you.

```{r}
data_directory <- here::here("materials", "data")

if (!dir.exists(data_directory)) {
  dir.create(data_directory)
}
```


## Yelp reviews

Used for sentiment classification. Contains 1,569,264 samples from the 
[Yelp Dataset Challenge 2015](https://www.yelp.com/dataset/challenge). This 
subset has 280,000 training samples and 19,000 test samples in each polarity.

```{r}
yelp_data <- file.path(data_directory, "yelp_review_polarity_csv")
if (!dir.exists(yelp_data)) {
  url <- "http://s3.amazonaws.com/fast-ai-nlp/yelp_review_polarity_csv.tgz"
  download.file(url, destfile = "materials/data/tmp.tar.gz")
  untar("materials/data/tmp.tar.gz", exdir = "materials/data/")
  invisible(file.remove("materials/data/tmp.tar.gz"))
}
```

