---
title: "Mini-project: Ames -- Regression to predict Ames, IA Home Sales Prices"
output:
  html_notebook:
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
ggplot2::theme_set(ggplot2::theme_minimal())
```

In this case study, our objective is to predict the sales price of a home. This 
is a _regression_ problem since the goal is to predict any real number across
some spectrum (\$119,201, \$168,594, \$301,446, etc). To predict the sales 
price, we will use numeric and categorical features of the home.

As you proceed, you'll work through the steps we discussed in the last module:

1. Prepare data
2. Balance batch size with a default learning rate
3. Tune the adaptive learning rate optimizer
4. Add callbacks to control training
5. Explore model capacity
6. Regularize overfitting
7. Repeat steps 1-6
8. Evaluate final model results

# Package Requirements

```{r load-pkgs}
library(keras)     # for deep learning
library(testthat)  # unit testing
library(tidyverse) # for dplyr, ggplot2, etc.
library(rsample)   # for data splitting
library(recipes)   # for feature engineering
```

# Step 0: Our Data

## The Ames housing dataset

For this case study we will use the [Ames housing dataset](http://jse.amstat.org/v19n3/decock.pdf) 
provided by the __AmesHousing__ package.

```{r get-data}
ames <- AmesHousing::make_ames()
dim(ames)
```

## Understanding our data

This data has been partially cleaned up and has no missing data:

```{r}
sum(is.na(ames))
```

But this tabular data is a combination of numeric and categorical data that we
need to address.

```{r ames-structure}
str(ames)
```

The numeric variables are on different scales. For example:

```{r numeric-ranges}
ames %>%
  select(Lot_Area, Lot_Frontage, Year_Built, Gr_Liv_Area, Garage_Cars, Mo_Sold) %>%
  gather(feature, value) %>%
  ggplot(aes(feature, value)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::comma)
```

There are categorical features that could be ordered:

```{r numeric-categories}
ames %>%
  select(matches("(Qual|Cond|QC|Qu)$")) %>%
  str()
```

And some of the categorical features have many levels:

```{r}
ames %>%
  select_if(~ is.factor(.) & length(levels(.)) > 8) %>%
  glimpse()
```

Consequently, our first challenge is transforming this dataset into numeric
tensors that our model can use.

# Step 1: Prep the Data

## Create train & test splits

One of the first things we want to do is create a train and test set as you
probably noticed that we do not have a train and test set similar to how MNIST 
was already set up for us. We can use the __rsample__ package to create our
train and test datasets.

__Note__: This will randomly select the 70/30 split so we are randomizing our
data with this process.

```{r}
set.seed(123)
ames_split <- initial_split(ames, prop = 0.7)
ames_train <- analysis(ames_split)
ames_test <- assessment(ames_split)

dim(ames_train)
dim(ames_test)
```

## Vectorize and scaling

All inputs and response values in a neural network must be tensors of either 
floating-point or integer data. Moreover, our feature values should not be
relatively large compared to the randomized initial weights _and_ all our 
features should take values in roughly the same range.

Consequently, we need to ___vectorize___ our data into a format conducive to neural 
networks [ℹ️](http://bit.ly/dl-02#3). For this data set, we'll transform our
data by:

1. removing any zero-variance (or near zero-variance) features
2. condensing unique levels of categorical features to "other"
3. ordinal encoding the quality features
4. normalize numeric feature distributions
5. standardizing numeric features to mean = 0, std dev = 1
6. one-hot encoding remaining categorical features

__Note__: we're using the recipes package (https://tidymodels.github.io/recipes)

```{r}
blueprint <- recipe(Sale_Price ~ ., data = ames_train) %>%
  step_nzv(all_nominal()) %>%                                       # step #1
  step_other(all_nominal(), threshold = .01, other = "other") %>%   # step #2
  step_integer(matches("(Qual|Cond|QC|Qu)$")) %>%                   # step #3
  step_YeoJohnson(all_numeric(), -all_outcomes()) %>%               # step #4
  step_center(all_numeric(), -all_outcomes()) %>%                   # step #5
  step_scale(all_numeric(), -all_outcomes()) %>%                    # step #5
  step_dummy(all_nominal(), -all_outcomes(), one_hot = TRUE)        # step #6

blueprint
```

This next step computes any relavent information (mean and std deviation of
numeric features, names of one-hot encoded features) on the training data so
there is no information leakage from the test data.

```{r}
prepare <- prep(blueprint, training = ames_train)
prepare
```

We can now vectorize our training and test data. If you scroll through the data
you will notice that all features are now numeric and are either 0/1 (one hot
encoded features) or have mean 0 and generally range between -3 and 3.

```{r}
baked_train <- bake(prepare, new_data = ames_train)
baked_test <- bake(prepare, new_data = ames_test)

# unit testing to ensure all columns are numeric
expect_equal(map_lgl(baked_train, ~ !is.numeric(.)) %>% sum(), 0)
expect_equal(map_lgl(baked_test, ~ !is.numeric(.)) %>% sum(), 0)

baked_train
```

Lastly, we need to create the final feature and response objects for train and 
test data. Since __keras__ and __tensorflow__ require our features & labels to be 
seperate objects we need to separate them. In doing so, our features need to be 
a 2D tensor which is why we apply `as.matrix` and our response needs to be a 
vector which is why we apply `pull`.

```{r}
x_train <- select(baked_train, -Sale_Price) %>% as.matrix()
y_train <- baked_train %>% pull(Sale_Price)

x_test <- select(baked_test, -Sale_Price) %>% as.matrix()
y_test <- baked_test %>% pull(Sale_Price)

# unit testing to x & y tensors have same number of observations
expect_equal(nrow(x_train), length(y_train))
expect_equal(nrow(x_test), length(y_test))
```

Our final feature set now has 188 input variables:

```{r}
dim(x_train)
dim(x_test)
```

# Step 2: Balance batch size with a default learning rate

To get started, let's build a simple model with...

- a single layer model with 128 units in the hidden layer. We have 188 features
  and 1 response node so a good starting point is mean(c(188, 1)) and then round
  up to the nearest value in the $2^s$ range (i.e. 32, 64, 128, 256, 512).
- a basic SGD optimizer
- use a mean square logarithmic error ("msle")
- also track the mean absolute error metric ("mae")
- 20% validation split

Now, start with the default batch size of 32 and then compare with smaller
values (i.e. 16) and larger values (i.e. 128). You're looking to balance the
progression of the loss learning curve and the training spead.

__Comment__: The default batch size of 32 performs pretty well in this case but
you could've easily have choosen lower (8 or 16) or higher (64, 128) without
negative impacts. We can see that the loss is still trending downward so we
should have lots of room for improvement.

```{r}
n_feat <- ncol(x_train)

model <- keras_model_sequential() %>% 
  layer_dense(units = 128, activation = "relu", input_shape = ncol(x_train)) %>%
  layer_dense(units = 1)

model %>% compile(
    optimizer = "sgd",
    loss = "msle",
    metrics = "mae"
  )

history <- model %>% fit(
  x_train,
  y_train,
  batch_size = 32,
  validation_split = 0.2
)
```

```{r}
history
```

```{r}
plot(history)
```

# Step 3: Tune the adaptive learning rate optimizer

Now go head and start assessing different adaptive learning rates such as:

- SGD+momentum
- RMSprop
- Adamp

Try a variety of learning rates. Recall that we typically start assessing rates
on a logarithmic scale (i.e. 0.1, 0.01, ..., 0.0001).

__Comment__: The default learning rates on the common adaptive learning rate
optimizers show a slow progression down the loss curve so we can afford to use
a larger learning rate. I found that the RMSprop tended to provide the best
results at this point.

```{r]}
model <- keras_model_sequential() %>% 
  layer_dense(units = 128, activation = "relu", input_shape = ncol(x_train)) %>%
  layer_dense(units = 1)

model %>% compile(
    optimizer = optimizer_rmsprop(lr = 0.1),
    loss = "msle",
    metrics = "mae"
  )

history <- model %>% fit(
  x_train,
  y_train,
  batch_size = 32,
  validation_split = 0.2
)
```

```{r}
history
```

```{r}
plot(history)
```

# Step 4: Add callbacks to control training

Add the following callbacks and see if your performance improves:

- early stopping with `patience = 3` and `min_delta = 0.00001`
- learning rate reduction upon a plateau with `patience = 1`

__Comment__: Adding early stopping improves performance because we can increase
the epochs but stop when necessary. Most of my optimal models were stopping at
around 15 epochs at this point. Also, adding `callback_reduce_lr_on_plateau()`
also improves. I plot the learning rates by epoch below and we can see that
they reduce multiple times which allows our model to eek out a little more
performance improvements.

```{r]}
model <- keras_model_sequential() %>% 
  layer_dense(units = 128, activation = "relu", input_shape = ncol(x_train)) %>%
  layer_dense(units = 1)

model %>% compile(
    optimizer = optimizer_rmsprop(lr = 0.1),
    loss = "msle",
    metrics = "mae"
  )

history <- model %>% fit(
  x_train,
  y_train,
  batch_size = 32,
  epochs = 30,
  validation_split = 0.2,
  callbacks = list(
    callback_early_stopping(patience = 3, min_delta = 0.00001),
    callback_reduce_lr_on_plateau(patience = 1)
  )
)
```

```{r}
history
```

```{r}
plot(history)
```

Plotting the learning rate shows that it reduced multiple times during training:

```{r}
plot(history$metrics$lr)
```


# Step 5: Explore model capacity

Now start to explore different widths and depths to your model.

- Assess a single layer with 128, 256, 512, and 1024 nodes
- Assess 1, 2, and 3 hidden layers

__Comment__: I follow the same approach we used in the previous module to 
assess combinations of different nodes and hidden layers. I used the tensorboard
callback to save my model runs and analyze them.

```{r}
train_model <- function(n_units, n_layers, log_to) {
  
  # Create a model with a single hidden input layer
  model <- keras_model_sequential() %>%
    layer_dense(units = n_units, activation = "relu", input_shape = n_feat)
  
  # Add additional hidden layers based on input
  if (n_layers > 1) {
    for (i in seq_along(n_layers - 1)) {
      model %>% layer_dense(units = n_units, activation = "relu")
    }
  }
  
  # Add final output layer
  model %>% layer_dense(units = 1)
  
  # compile model
  model %>% compile(
    optimizer = optimizer_rmsprop(lr = 0.1),
    loss = "msle",
    metrics = "mae"
  )
  
  # train model and store results with callback_tensorboard()
  history <- model %>% fit(
  x_train,
  y_train,
  batch_size = 32,
  epochs = 30,
  validation_split = 0.2,
  callbacks = list(
    callback_early_stopping(patience = 3, min_delta = 0.00001),
    callback_reduce_lr_on_plateau(patience = 1),
    callback_tensorboard(log_dir = log_to)
  ),
  verbose = FALSE
  )
  
  return(history)
}
```

```{r}
grid <- expand_grid(
  units = c(128, 256, 512, 1024),
  layers = c(1:3)
) %>%
  mutate(id = paste0("mlp_", layers, "_layers_", units, "_units"))
grid
```

The initial results don't show any glaring trends. All our models have loss
scores ranging from 0.0135-0.015.

```{r}
for (row in seq_len(nrow(grid))) {
  # get parameters
  units <- grid[[row, "units"]]
  layers <- grid[[row, "layers"]]
  file_path <- paste0("ames/", grid[[row, "id"]])
  
  # provide status update
  cat(layers, "hidden layer(s) with", units, "neurons: ")
  
  # train model
  m <- train_model(n_units = units, n_layers = layers, log_to = file_path)
  min_loss <- min(m$metrics$val_loss, na.rm = TRUE)
  
  # update status with loss
  cat(min_loss, "\n", append = TRUE)
}
```

Looking at the tensorboard shows that, really, any of the models are decent
choices as they all have relatively similar results, low variance which means
they all are stable models, they all have minimal overfitting, and compute time
is definitely not a problem.

```{r}
tensorboard("ames")
```

__Comment__: After a little more experimenting I found a funnel shaped approached tended to
produce a little more improvement in performance:

```{r}
model <- keras_model_sequential() %>% 
  layer_dense(units = 1024, activation = "relu", input_shape = n_feat) %>%
  layer_dense(units = 512, activation = "relu") %>%
  layer_dense(units = 256, activation = "relu") %>%
  layer_dense(units = 1)

model %>% compile(
    optimizer = optimizer_rmsprop(lr = 0.1),
    loss = "msle",
    metrics = "mae"
  )

history <- model %>% fit(
  x_train,
  y_train,
  batch_size = 32,
  epochs = 30,
  validation_split = 0.2,
  callbacks = list(
    callback_early_stopping(patience = 3, min_delta = 0.00001),
    callback_reduce_lr_on_plateau(patience = 1)
  )
)
```

```{r}
history
```

```{r}
plot(history) + scale_y_log10()
```

# Step 6: Regularize overfitting

If your model is overfitting, try to add...

- weight decay (i.e. `kernel_regularizer = regularizer_l2(l = xxx)`). Remember,
  we typically start by assessing values on logarithmic scale [0.1, 0.00001].
- dropout (`layer_dropout()`) between each layer. Remember, dropout rates
  typically range from 20-50%.
  
__Comment__: Pretty much any weight regularizer hurt model performance. In this
case, since our validation loss has minimal overfitting we can probably disregard
any additional regularization.
  
```{r}
model <- keras_model_sequential() %>% 
  layer_dense(units = 1024, activation = "relu", input_shape = n_feat) %>%
  layer_dropout(0.2) %>%
  layer_dense(units = 512, activation = "relu") %>%
  layer_dropout(0.2) %>%
  layer_dense(units = 256, activation = "relu") %>%
  layer_dropout(0.2) %>%
  layer_dense(units = 1)

model %>% compile(
    optimizer = optimizer_rmsprop(lr = 0.1),
    loss = "msle",
    metrics = "mae"
  )

history <- model %>% fit(
  x_train,
  y_train,
  batch_size = 32,
  epochs = 30,
  validation_split = 0.2,
  callbacks = list(
    callback_early_stopping(patience = 3, min_delta = 0.00001),
    callback_reduce_lr_on_plateau(patience = 1)
  )
)
```

```{r}
history
```

# Step 7: Repeat steps 1-6

As this point we could repeat the process and...

1. Prepare data
   - try to find additional data to add
   - try new feature engineering approaches
2. Balance batch size with a default learning rate
   - reassess batch size
3. Tune the adaptive learning rate optimizer
   - fine tune our learning rate
   - see if the current optimizer still outperforms others
4. Add callbacks to control training
   - maybe assess more sophisticated learning rate schedulers (i.e. cyclical
     learning rates)
5. Explore model capacity
   - after some tweaks we may want to reassess model capacity combinations
6. Regularize overfitting

[🏠](https://github.com/rstudio-conf-2020/dl-keras-tf)