---
title: "NLP: Transfer learning for Amazon review word embeddings"
output:
  html_notebook:
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

This project is designed to test your current knowledge on applying word
embeddings to the [Amazon Fine Foods reviews](https://snap.stanford.edu/data/web-FineFoods.html) 
dataset available through Stanford. This dataset contains 568,454 reviews on
74,258 products.

Your goal is to develop a word embedding model to accurately predict how 
__helpful__ a review will be. I supply code to help you get the data
imported and prepped so that you can focus on the modeling aspect.

___Good luck!___

# Requirements

```{r, message=FALSE, warning=FALSE}
library(keras)     # provides deep learning procedures
library(tidyverse) # provides basic data wrangling and visualization
library(glue)      # provides efficient print statements
library(testthat)  # provides unit testing
```

# Data importing

The [finefoods.txt.gz](https://snap.stanford.edu/data/finefoods.txt.gz) file has
already been downloaded and unzipped for you. All reviews are contained in a
single .txt file.

```{r}
amazon_reviews <- here::here("materials", "data", "amazon-food", "finefoods.txt")
reviews <- read_lines(amazon_reviews)
```

Each review consists of 8 items and each item is on its own line. The following
shows all information collected for the first review.

```{r}
head(reviews, 8)
```

# Verify we properly imported

Based on the data's [website](https://snap.stanford.edu/data/web-FineFoods.html),
we should have the following:

- Number of reviews: 568,454
- Number of products: 74,258
- Number of users: 256,059

```{r}
review_text <- reviews[str_detect(reviews, "review/text:")]
products <- reviews[str_detect(reviews, "product/productId:")]
users <- reviews[str_detect(reviews, "review/userId:")]

n_reviews <- length(review_text)
n_products <- n_distinct(products)
n_users <- n_distinct(users)

# Verify our imported data aligns with data codebook
expect_equal(n_reviews, 568454)
expect_equal(n_products, 74258)
expect_equal(n_users, 256059)
```


# Extracting key parts of the data

There are two main parts of these reviews that we need for our modeling purpose:

1. The review text
2. The fraction of users who found the review helpful

## Getting our text

Let's extract the text

```{r}
text <- review_text %>%
  str_replace("review/text:", "") %>%
  iconv(to = "UTF-8") %>%
  str_trim()

expect_equal(length(text), n_reviews)

text[1]
```

## Getting our labels

Now let's extract our helpfulness information. This represents the fraction of
users who found the review helpful for a given product.

```{r}
helpfulness_info <- reviews[str_detect(reviews, "review/helpfulness:")] %>%
  str_extract("\\d.*")

expect_equal(length(helpfulness_info), n_reviews)

head(helpfulness_info)
```

Let's separate this information into the number of reviews (denominator) and
the number of user who found the review helpful (numerator).

```{r}
num_reviews <- str_replace(helpfulness_info, "^.*\\/", "") %>% as.integer()
helpfulness <- str_replace(helpfulness_info, "\\/.*$", "") %>% as.integer()
```

And we're only going to care about those products with 10+ reviews to try
minimize some of the noise.

```{r}
num_index <- num_reviews >= 10
num_reviews <- num_reviews[num_index]
helpfulness <- helpfulness[num_index]
text <- text[num_index]

# verify that the number of observations in each vector is equal
expect_equal(
  map_int(list(num_reviews, helpfulness, text), length) %>% n_distinct(),
  1
)

glue("There are {sum(num_index)} observations with 10 or more reviews.")
```

Our labels are going to be the fraction provided by helpfulness converted to a
percentage.

```{r}
labels <- helpfulness / num_reviews

expect_equal(length(labels), length(text))

range(labels)
```

We can look at a review that is considered very helpful...

```{r}
first_pos <- first(which(labels == 1))
text[first_pos]
```

versus a review that is considered very unhelpful.

```{r}
first_neg <- first(which(labels == 0))
text[first_neg]
```

Let's get a quick assessment of word usage across the reviews:

```{r}
text_df <- text %>%
  tibble(.name_repair = ~ "text") %>%
  mutate(text_length = str_trim(text) %>% str_count("\\w+"))

unique_words <- text_df %>%
  tidytext::unnest_tokens(word, text) %>%
  pull(word) %>%
  n_distinct()

avg_review_length <- median(text_df$text_length, na.rm = TRUE)
  
ggplot(text_df, aes(text_length)) +
  geom_histogram(bins = 100, fill = "grey70", color = "grey40") +
  geom_vline(xintercept = avg_review_length, color = "red", lty = "dashed") +
  scale_x_log10() +
  ggtitle(glue("Median review length is {avg_review_length}"),
          subtitle = glue("Total number of unique words is {unique_words}"))
```


# Explore Glove Embeddings

We can explore word embeddings that give us some context of the review language.

```{r}
# helper functions we'll use to explore word embeddings
source("helper_functions.R")

# clean up text and compute word embeddings
clean_text <- tolower(text) %>%
  str_replace_all(pattern = "[[:punct:] ]+", replacement = " ") %>%
  str_trim()

word_embeddings <- get_embeddings(clean_text)
```

Explore your own words!

```{r}
# find words with similar embeddings
get_similar_words("oil", word_embeddings)
```

# Prepare data

Our labels are already a tensor (vector) so we don't need to do any additional
prep.

```{r}
str(labels)
```

## Preprocessing hyperparameters

However, we need to preprocess our text. First, lets decide on two key parameters
to use when preprocessing our text:

1. number of most frequent words used (start with 20000)
2. the maximum length of our processed text (start with 200)

These are two hyperparameters you can come back to and change as hyperparameters.

```{r}
top_n_words <- 20000
max_len <- 200
```

## Preprocessing Feature text

Next, you need to create and apply a tokenizer to the text.

```{r}
tokenizer <- text_tokenizer(num_words = top_n_words) %>% 
  fit_text_tokenizer(text)

names(tokenizer)
```

Now, convert your text to a numerically encoded sequence.

```{r}
sequences <- texts_to_sequences(tokenizer, text)
```

```{r}
# The vectorized first instance:
sequences[[1]]
```

Run this code chunk to see how your text has been converted:

```{r} 
cat(crayon::blue("Original text:\n"))
text[[1]]

cat(crayon::blue("\nRevised text:\n"))
paste(unlist(tokenizer$index_word)[sequences[[1]]] , collapse = " ")
```

Last, we want to make sure our sequences (aka each processed review) is of equal
length.

```{r}
features <- pad_sequences(sequences, maxlen = max_len)

expect_equal(ncol(features), max_len)
```

Make sure that the number of observations in your features and labels are equal:

```{r}
expect_equal(nrow(features), length(labels))
```


# Model training

Before we train our model, let's go ahead and randomize our review data so that
our training and validation data properly represent a mixture of products and
users.

```{r}
set.seed(123)
index <- sample(1:nrow(features))
split_point <- floor(length(index) * .3)
train_index <- index[1:split_point]
valid_index <- index[(split_point + 1):length(index)]

expect_equal(length(train_index) + length(valid_index), length(index))

x_train <- features[train_index, ]
y_train <- labels[train_index]

x_valid <- features[valid_index, ]
y_valid <- labels[valid_index]
```

Ok, so before we train our model, let's get an understanding of a baseline
loss score that we want to beat. The easiest baseline is to just predict the
average of the training label for future observations.

```{r}
avg <- mean(y_train)
baseline_mse <- mean((y_valid - avg)^2)

cat("Simply predicting the average helpfulness score of", round(avg, 2),
    "for every review would give us a loss score of", round(baseline_mse, 3))
```

Ok, time to build your model architecture and compile it. Since this is a
regression problem but we want to keep our predicted values bounded between 0
and 1, I create a custom "clipped" MSE metric. This is not something I expected
you to do but wanted to illustrate here.

```{r}
model <- keras_model_sequential() %>%
  layer_embedding(input_dim = top_n_words, 
                  output_dim = 32,
                  input_length = max_len) %>% 
  layer_flatten() %>%
  layer_dense(units = 32, activation = "relu") %>%
  layer_dropout(0.5) %>%
  layer_dense(units = 1)

# create metric using backend tensor functions
metric_clipped_mse <- custom_metric("metric_capped_mse", function(y_true, y_pred) {
  y_pred <- k_clip(y_pred, 0, 1)
  k_mean(k_square(y_pred - y_true))
})

model %>% compile(
  optimizer = optimizer_rmsprop(lr = 0.0001),
  loss = metric_clipped_mse,
  metrics = c("mse", "mae")
)

summary(model)
```

Let's train our model:

```{r}
history <- model %>% fit(
  x_train, y_train,
  epochs = 50,
  batch_size = 128,
  validation_data = list(x_valid, y_valid),
  callbacks = list(
    callback_reduce_lr_on_plateau(patience = 3),
    callback_early_stopping(patience = 10, restore_best_weights = TRUE)
    )
)
```

In this example we see that our model's optimal validation loss was 0.057, which
is 33% lower than the baseline loss. 

```{r}
opt_mse <- min(history$metrics$val_loss)
glue("Baseline loss score: {round(baseline_mse, 3)}")
glue("Model loss score: {round(opt_mse, 3)}")
```

The following plots the actual "helpfulness" value for each observation (dots)
compared to our model's predicted value (red line) and the average predicted
value (blue). We see that our model is picking up some signal and is a better
representation than the average value.

```{r}
tibble(
  actual = y_valid,
  pred = model %>% predict(x_valid) %>% as.vector()
  ) %>%
  mutate(pred = case_when(
    pred < 0 ~ 0,
    pred > 1 ~ 1,
    TRUE ~ pred
  )) %>%
  arrange(pred) %>%
  mutate(id = row_number()) %>%
  ggplot(aes(x = id)) +
  geom_point(aes(y = actual), size = 1, alpha = 0.2) +
  geom_line(aes(y = pred), color = "red", size = 1) +
  geom_hline(yintercept = avg, lty = "dashed", color = "blue")
```

# How did I choose my parameters

You may be wondering how I chose my hyperparameters (i.e. embedding layer output
dimension, number of hidden layers and units, learning rate)? Well, unlike you
I had the luxery of time to run a grid search. You can check it out
[here](https://rstudio-conf-2020.github.io/dl-keras-tf/notebooks/amazon-embeddings-grid-search.nb.html). 

However, realize this is not necessarily an optimized solution so you could
continue improve upon it!

[🏠](https://github.com/rstudio-conf-2020/dl-keras-tf)