---
title: "Linear regression with stochastic gradient descent"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
ggplot2::theme_set(ggplot2::theme_minimal())
```

This notebook illustrates how one can learn parameters using stochastic
gradient descent. This will demonstrate the basic idea of how iterating over
forward and backward passes improves our loss function.

A couple things to note. First, this is note building a neural net from scratch;
rather, it is using stochasti gradient descent to build a simple linear
regression model which would be analogous to building a single perceptron.
Second, this notebook does not focus on efficient implementation of this
algorithm; rather, the code is designed to make the concept intuitive to the
reader.

# Requirements

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(glue)
```

# Creating our data

For simplicity we are going to use a very simple data set that is generated
based on a simple linear model without any error. Our data includes:

- 30 observations (n),
- with an intercept of 2 (b),
- and slope of 30 (a)

```{r}
n <- 30
b <- 30
a <- 2

set.seed(123)
(df <- tibble(
  x = runif(30, 0, 100), 
  y = b + a*x
  ))
```

The following plot shows how an ordinary least squares model aligns to the
underlying data.

```{r}
ggplot(df, aes(x, y)) +
  geom_smooth(method = "lm", se = FALSE, lty = "dashed", size = 0.75) +
  geom_point(size = 2) +
  ggtitle("Ordinary least squares model")
```


# Learning parameters with gradient descent

Our quest is to learn this linear relationship using gradient descent. Recall
that learning based on gradient descent includes the following steps
[ℹ️](http://bit.ly/dl-01#49):

1. Perform a forward pass to compute predicted values,
2. Compute the loss of our predictions,
3. Compute the gradient,
4. Update our weights,
5. Repeat until loss is sufficiently minimized.

## Forward pass

First, we start with some randomized values for our bias (intercept) and slope
(weight). 

```{r}
(bias <- runif(1))
(weight <- runif(1))
```

This notebook demonstrates stochastic gradient descent which means we are using
a batch size of 1. Later notebooks will demo batch sizes > 1. For each step that
we take we'll build a helper function. This first function simply grabs the
batch (aka row) of interest.

```{r}
get_batch <- function(.data, batch_id) {
  .data %>% 
    slice(batch_id)
}

df %>% get_batch(batch_id = 1)
```

The next step is to make a prediction with our randomized weights:

```{r}
make_prediction <- function(.data, bias, weight) {
  .data %>% mutate(pred = bias + weight * x)
}

df %>%
  get_batch(batch_id = 1) %>%
  make_prediction(bias, weight)
```

Now we can compute the error of this prediction which is just the squared error:

```{r}
compute_sq_error <- function(.data, actual, predicted) {
  .data %>% mutate(error = ({{actual}} - {{predicted}})^2)
}

df %>%
  get_batch(batch_id = 1) %>%
  make_prediction(bias, weight) %>%
  compute_sq_error(y, pred)
```

## Backward pass

Now that we have our error, we can use this information to update our weights.
To do so, we need to compute the derivative which we can use to update the
weights.

### Compute the derivative

There are two ways to compute the derivative. The first is more intuitive and is
known as finite differencing. This provides an _estimate_ of the derivative by
simply:

1. adding a small amount to the weight and bias, 
2. making a new prediction with these adjusted amounts,
3. computing the new error,
4. and using this information to compute the derivative

The following makes two new predictions; one where we add a small amount to our
weight and another where we add a small amount to our bias. We then compute the
error of these predictions.

We can see that by increasing the weight and the bias our error gets smaller.

Now we can compute the derivative. Recall that the derivative is just how much
y changes divided by how x changes. Here, we make a small adjustment of 0.01 to
our bias and weight and in both cases, this adjustment causes a small decrease
to our error.

```{r}
compute_est_derivatives<- function(.data, bias, weight, adj_amt) {
  .data %>% 
    # first compute difference in errors with adjusted bias and weight values
    mutate(
      pred_w_adj_bias = (bias + adj_amt) + weight * x,
      pred_w_adj_weight = bias + (weight + adj_amt) * x,
      error_adj_bias = (y - pred_w_adj_bias)^2,
      error_adj_weight = (y - pred_w_adj_weight)^2
      ) %>%
    # now compute estimated derivative based on adjust predictions
    mutate(
      est_deriv_bias = (error_adj_bias - error) / adj_amt,
      est_deriv_wt = (error_adj_weight - error) / adj_amt,
    ) %>%
    select(-contains("pred_w_adj"))
}
  
df %>%
  get_batch(batch_id = 1) %>%
  make_prediction(bias, weight) %>%
  compute_sq_error(y, pred) %>%
  compute_est_derivatives(bias, weight, adj_amt = 0.01)
```

The other approach to computing derivatives is to actually compute the
derivatives and integrals. With this example, we can do this easily because our
underlying function is not complicated.

Considering our underlying function is $y = bias + 2x$ then the:

* first derivitive = $\frac{\delta \text{ error}}{\delta \text{ weight}} = 2 * (\hat{y} - y)$
* second derivitive = $\frac{de}{da} = x * \frac{\delta \text{ error}}{\delta \text{ weight}}$

There are some R packages that provide auto-differentiation; however, you will
never need to manually compute derivatives so we'll just go off our a priori
knowledge of the underlying function. Yes this is cheating a bit but right now
we're just trying to gain intuition for the gradient descent process.

We can see that the estimated derivatives (`est_deriv_bias` & `est_deriv_wt`)
are good approximates for the actual derived derivatives (`deriv_bias` &
`deriv_wt`).

```{r}
compute_derivatives <- function(.data, predicted, actual, x_values) {
  .data %>% 
    mutate(
      deriv_bias = 2 * ({{predicted}} - {{actual}}),
      deriv_wt = deriv_bias * {{x_values}}
    )
}

df %>%
  get_batch(batch_id = 1) %>%
  make_prediction(bias, weight) %>%
  compute_sq_error(y, pred) %>%
  compute_est_derivatives(bias, weight, adj_amt = .01) %>%
  compute_derivatives(pred, y, x)
```

### Updating our weights

To update our model we first need to decide how large of a step we want to take
down the gradient descent.This is called the learning rate. New values for our
bias and weights are simply the original values minus the derivative times the
learning rate. 

Here, we use a learning rate of `0.0001`. You can see that both the updated bias
and weight (`new_bias = 0.975` & `new_wt = 1.25`) have increased in a positive
direction compared to the original values (original bias = 0.96 & original
weight = 0.90).

```{r}
update_value <- function(current_value, derivative, learning_rate) {
  current_value - (derivative * learning_rate)
}

df %>%
  get_batch(batch_id = 1) %>%
  make_prediction(bias, weight) %>%
  compute_sq_error(y, pred) %>%
  compute_est_derivatives(bias, weight, adj_amt = .01) %>%
  compute_derivatives(pred, y, x) %>%
  mutate(new_bias = update_value(bias, deriv_bias, 0.0001),
         new_wt = update_value(weight, deriv_wt, 0.0001))
```

## Cycling through an epoch

Recall that after we update our weights, we execute a new batch following the
same procedure. We do this for every batch until we have worked through the
entire data set and that is considered one epoch. Let's do this procedure for
our data. 

First, let's bring all the steps we just went through together into a single
function that does a forward and backward pass on a single batch.

```{r}
compute_batch_update <- function(.data, bias, weight, learning_rate, batch) {
  .data %>%
    get_batch(batch_id = batch) %>%
    make_prediction(bias, weight) %>%
    compute_sq_error(y, pred) %>%
    compute_est_derivatives(bias, weight, adj_amt = .01) %>%
    compute_derivatives(pred, y, x) %>%
    mutate(new_bias = update_value(bias, deriv_bias, learning_rate),
           new_wt = update_value(weight, deriv_wt, learning_rate))
}
```


```{r}
compute_batch_update(df, initial_bias, initial_weight, learning_rate = 0.0001, batch = 1)
```

We can now create a function that iterates through each batch for the entire
data set (1 epoch):

```{r}
train_epoch <- function(.data, bias, weight, learning_rate) {
  epoch_results <- data.frame()
  for (batch in seq_len(nrow(.data))) {
    batch_results <- compute_batch_update(.data, bias, weight, learning_rate, batch)
    epoch_results <- rbind(epoch_results, batch_results)
    bias <- batch_results[["new_bias"]]
    weight <- batch_results[["new_wt"]]
    }
  epoch_results
  }
```

If we apply this function we'll see that the output is our original data frame
with the deriviatives and updated bias and weights computed based on each
observation. 

```{r}
train_epoch(df, bias, weight, learning_rate = 0.0001)
```

## Cycling through multiple epochs

Now let's create one last function wich will apply our functions for multiple
epochs. This function also provides a new data frame that shows our loss score
(MSE) for each epoch along with the original and updated biases & weights.

```{r}
train_model <- function(.data, bias, weight, learning_rate, epochs) {
  results <- data.frame()
  for (epoch in seq_len(epochs)) {
    epoch_results <- train_epoch(.data, bias, weight, learning_rate) %>%
      mutate(
        bias = bias, 
        weight = weight,
        epoch = epoch,
        mse = sqrt(sum(error))
        ) %>%
      slice(n()) %>%
      select(epoch, bias, weight, mse, new_bias, new_weight = new_wt)
    
    results <- rbind(results, epoch_results)
    
    bias <- epoch_results[["new_bias"]]
    weight <- epoch_results[["new_weight"]]
  }
  results
}
```

Since we are working with a small learning rate and SGD does not include any
momentum, we'll train this over many epochs. Looking at the last 10 rows in our
output we see that the `new_bias` and `new_weight` nearly equal the underlying
bias (30) and weight (2) that our data was generated with.

```{r}
history <- train_model(df, bias, weight, learning_rate = 0.0001, epochs = 4000)
tail(history, 10)
```

The following plot shows the progression down the loss learning curve for each
epoch; practically reach 0 towards the end.

```{r}
ggplot(history, aes(epoch, mse)) +
  geom_point()
```

The following grabs the results from just some of our epochs (epoch 100, 200, ...,
4000) and plots the predicted values based on the estimated bias and weight
parameters. You can see how the final estimated bias and weight parameters
converge to the same prediction plot illustrated earlier using ordinary least
squares.

```{r}
subset_history <- history %>% filter((epoch %% 100) == 0)

prediction_change <- data.frame()

for (row in seq_len(nrow(subset_history))) {
  bias <- subset_history[[row, "new_bias"]]
  weight <- subset_history[[row, "new_weight"]]
  epoch <- subset_history[[row, "epoch"]]
  y_hat <- make_prediction(df, bias, weight) %>%
    mutate(epoch = epoch)
  prediction_change <- rbind(prediction_change, y_hat)
}

ggplot(prediction_change, aes(x, y)) +
  geom_line(aes(y = pred, group = factor(epoch)), alpha = 0.3) +
  geom_point(color = "red")
```

